### 2014-12-26 mcmc.tex
methods content

MCMC methods are proven to be correct ``in the limit''; that is, when
an infinite number of samples have been taken.
When the function $f(\pars)$ has challenging properties, the results
can approach this limit very slowly (or not at all\note{HOGG: discuss
  situations in which MCMC will not work---related to support.}).
For this reason, in the finite duration of a scientific project, most
users of MCMC have one of two (related) problems:
Either it is taking too long---they have take too many steps or
execute too many calls of the function $f(\pars)$---or it is not
exploring the full parameter space---there are local optima (or bad
local geometry around those optima) ``trapping'' the algorithm.

There is no general, problem-independent solution for either of these
problems.
Indeed, there are many different strategies that work well or badly
for different functions $f(\pars)$.
In this \sectionname, we will try to guide the reader towards methods
that might help in different circumstances.
We will not give a full description of all the MCMC algorithms out
there, but just provide some field notes and pointers to references.
New MCMC methods are being developed all the time, so a reader with a
serious problem would do well to consult with the applied-mathematics
or statistics literature.\note{HOGG: Your best interface to these
  literatures is via a colleague.  Break down those walls!}
The different methods to which we refer here are (in general) very
different in their difficulty of implementation.
We won't help with that either, except to say that there are more and
more open-source or distributed packages every year.
We will try to name some current implementations in the notes, but any
list compiled today will be incomplete tomorrow.
The spaces of MCMC methods and MCMC method implementations are both
growing rapidly, in part because MCMC has become such a core
technology in the empirical sciences.

@@

...Gibbs!  Every parameter is differently difficult to sample in?

...What, briefly, is the stretch move, and why is it useful?  Every parameter treated identically---almost opposite to gibbs.

...What, briefly, is importance sampling?

...What, briefly, is nested sampling, and why is it good?

...What, briefly, is tempering?

...What, briefly, is hamiltonian, and why is it good?

...What, briefly, is reversible-jump, and why is it useful?

>>>>

[SHOULD THIS SECTION MOVE TO AFTER THE DEFINITION OF AUTOCORRELATION TIME AND DISCUSSION OF CONVERGENCE?]

MCMC methods are proven to be correct ``in the limit''; that is, when
an infinite number of samples have been taken.
When the function $f(\pars)$ has challenging properties, the results
can approach this limit very slowly (or not at all\note{HOGG: discuss
  situations in which MCMC will not work---related to support.}).
For this reason, in the finite duration of a scientific project, most
users of MCMC have one of two (related) problems:
Either \textsl{(1)}~it is taking too long---it is taking too many steps or
executing too many calls of the function $f(\pars)$ to get independent samples---or \textsl{(2)}~it is not
exploring the full parameter space---there are local optima (or bad
local geometry around those optima) ``trapping'' the algorithm.

There is no general, problem-independent solution for either of these
problems.
Indeed, there are many different strategies that work well or badly
for different functions $f(\pars)$.
In this \sectionname, we will try to guide the reader towards methods
that might help in different circumstances.
We will not give a full description of all the MCMC algorithms out
there, but just provide some field notes and pointers to references.
New MCMC methods are being developed all the time, so a reader with a
serious problem would do well to consult with the applied-mathematics
or statistics literature.\note{HOGG: Your best interface to these
  literatures is a colleague in another department.  Break down those walls!}
The different methods to which we refer here are (in general) very
different in their difficulty of implementation.
We won't help with that either, except to say that there are more and
more open-source or distributed packages every year.
We will try to name some current implementations in the notes, but any
list compiled today will be incomplete tomorrow.
The spaces of MCMC methods and MCMC method implementations are both
growing rapidly, in part because MCMC has become such a core
technology in the empirical sciences.

...Note a good general reference here?  MacKay?

\paragraph{Ensembles:}
As we discuss more below, one of the critical challenges in the use of
MCMC is tuning the proposal pdf $q(\pars'\given\pars)$.
In problems in which all of the parameters (all of the components or entries of $\pars$) are somehow ``equivalent'',
or the parameters can be seen as components of a vector in a vector
space, there are \emph{ensemble methods} that make use of not just one
random-walker but instead many walkers to automatically generate a properly tuned
proposal distribution.
These methods make use of the distribution of a set of independent
walkers in the parameter space ($\pars$-space) to gauge the typical step-sizes and
directions at which new proposals should be made, obviating or greatly
reducing the tuning requirements.
We are co-authors on one of these methods, \project{emcee}\note{HOGG:
  ref emcee}, which is popular in the astrophysical sciences.
Because ensemble methods are updating independent MCMC chains, they
are often easy to parallelize to make use of multiple cores.

\paragraph{Gibbs:}
In many problems, the parameters---the components or entries of the
blob $\pars$---are not equivalent at all.
Some are global parameters, which touch or have an effect on every
sub-part of the data blob $\data$, while others are local parameteres,
which only touch a small subset of the data.
Or, in another use case, some parameters are linear parameters
that---at fixed values of the other parameters---will have a Gaussian
likelihood or posterior pdf.
In these problems, there are some directions in parameter space that
are hard to move in (they require complete re-calculation of the full
non-linear function $f(\pars)$, and other directions that are very
easy to move in (they require only partial re-calculations or the
sampling could even be analytic and exact).
For problems with these structures, Gibbs samplers are ideal, and
there are good implementations.\note{HOGG cite JAGS}

A trivial kind of Gibbs sampling was mentioned above in the
discussion of M--H MCMC:
Even when running vanilla M--H MCMC, it is often useful to update only
one parameter at a time; that is, do only axis-aligned moves in the
$\pars$-space, cycling through axes as you go from step to step.
This helps with tuning and diagnosis, but it also permits a good
implementation to capitalize on simpler calculations for the simpler
parameter directions.

Gibbs sampling can be very good for exploiting multiprocessing:
Global parameter updates might require a serial re-calculation of the
full likelihood, but local parameter updates can be done in parallel,
with each local parameter working on its local bit of data all at the
same time.
This kind of structure is common in hierarchical inference, where
local parameters touch only individual objects in some population,
say, and global parameters are parameters \emph{of} that population as
a whole.
Parallelization is built in to some Gibbs implementations.\note{HOGG}

\paragraph{Hamiltonian:}
[HOGG: THIS HAS SOME OTHER NAME; WHAT?]
In many cases of interest, the function $f(\pars)$ can be analytically
differentiated, such that it is possible to quickly execute a gradient
$\dd f/\dd\pars$ evaluation.
Indeed, in the modern world of computing, there are even
auto-differentiation systems that deliver automatically code that
produces the gradient of any function you can write.\note{HOGG cite
  autodiff systems}
There is a class of MCMC methods that make use of gradient information
to effectively inform the proposal distribution and thus speed sampling.
... HOGG more here!

@@

\paragraph{Importance:}

\paragraph{Tempering:}

\paragraph{Nested:}

\paragraph{Reversible-jump:}
Finally, there are some problems that \emph{cannot be done} without
additional proposal technology...
